---
title: "SLMP GLMER model selection example"
author: "Marleen Haupt"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: inline
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
  
```{r libraries, include=FALSE}
## load previously installed packages into library 
library(lme4) #for mixed effect models
library(tidyverse)  #plotting, tabulating, etc.
library(markdown)
library(effects)
library(lattice)
library(reshape)
library(brms) #Bayesian analysis
library(performance) #looic
library(kableExtra) #for table formatting
library(psych)
library(car)
library(sjmisc) #for scaling all variables in a data set
library(dplyr)
```

# Background

The OCS-Plus is a tablet-based screening tool for cognitive impairment. In this research project, we are assessing healthy older participants and patients with amnestic mild cognitive impairment (aMCI, diagnosis based on neuropsychological testing) with this tablet test. The aim of the following analysis is to find out which subtests contribute most to the diagnosis of healthy (0) or aMCI (1). Therefore, this script uses all OCS-Plus subtests (18 in total) as the input for a logistic regression. Two of these subtests are continuous variables; the other subtest scores are counts/number of incidences so they are ordered but not truly continuous. 

# Overarching questions

1) Model selection approach: What is the best model selection approach if I have so many subtests? Do I use forward or backward model selection or a completely different approach?

2) Rescaling: Should I rescale my variables and, if yes, which ones? How do I then interpret the output and do I have to unstandardize my estimates?

3) Convergence problems: Why do I have convergence prpblems with the default Laplace approximation (nAGQ=1) and not when I set nAGQ = 0 and the integral is 'ignored' if I include all subtests? Is nAGQ=0 a valid setting or am I losing too much of the randomness structure? When I compare the models yielded by the forward selection process, one model has the best BIC even though it did not converge - how can I make sense of that? Can I interpret output value of a model that did not converge?

# Data preparation

```{r, include=TRUE, warning=FALSE}
## set your working directory
setwd("C:/Users/Neelram/Desktop/PhD/Courses/PotsdamSummerSchool")

## load in data
OCSMCI <- read.table('SLMPdata_MarleenHaupt.txt',header=T, sep="\t")

## outcome variable
OCSMCI$group <- as.factor(OCSMCI$group) #1 aMCI, 0 healthy controls
OCSMCI$ID <- as.factor(OCSMCI$ID)

## set contrasts
contrasts(OCSMCI$group)
contrasts(OCSMCI$group)=c(-1,1)

```

# GLMs including all subtests 

```{r, include=TRUE, warning=FALSE}

# all subtests
glm1 <- glmer(group ~ 1 + PictureNaming + Orientation + Semantics + Encoding1 + Encoding2 + DelayedRecall + DelayedRecallRecognition + EpisodicRecognition + TrailsProcessingSpeed + TrailsExecScore + RuleFinding + RulesLearned + FigureCopy + FigureRecall + Cancellation + Cancellation_FalsPos + InvisibleCancellation + InvisibleCancellation_Revisits + (1|ID),data=OCSMCI, family='binomial')

summary(glm1) #model does not converge, only two predictors have z values > 1.96, four more have z value > 1.6

# all subtests, bobyqa optimizer
glm2 <- glmer(group ~ 1 + PictureNaming + Orientation + Semantics + Encoding1 + Encoding2 + DelayedRecall + DelayedRecallRecognition + EpisodicRecognition + TrailsProcessingSpeed + TrailsExecScore + RuleFinding + RulesLearned + FigureCopy + FigureRecall + Cancellation + Cancellation_FalsPos + InvisibleCancellation + InvisibleCancellation_Revisits + (1|ID),data=OCSMCI, family='binomial', control=glmerControl(optimizer="bobyqa"))

summary(glm2) #model does not converge, no significant results with bobyqa 

# all subtests, nAGQ=0
glm3 <- glmer(group ~ 1 + PictureNaming + Orientation + Semantics + Encoding1 + Encoding2 + DelayedRecall + DelayedRecallRecognition + EpisodicRecognition + TrailsProcessingSpeed + TrailsExecScore + RuleFinding + RulesLearned + FigureCopy + FigureRecall + Cancellation + Cancellation_FalsPos + InvisibleCancellation + InvisibleCancellation_Revisits + (1|ID),data=OCSMCI, family='binomial', nAGQ=0)

summary(glm3) #model converges

# all subtests, no random effects (using glm instead of glmer)
glm4 <- glm(group ~ 1 + PictureNaming + Orientation + Semantics + Encoding1 + Encoding2 + DelayedRecall + DelayedRecallRecognition + EpisodicRecognition + TrailsProcessingSpeed + TrailsExecScore + RuleFinding + RulesLearned + FigureCopy + FigureRecall + Cancellation + Cancellation_FalsPos + InvisibleCancellation + InvisibleCancellation_Revisits,data=OCSMCI, family='binomial')

summary(glm4) #model converges, highly similar resutls to glm3

```

## Summary

The GLMs including all subtests (which would be the starting point for backward selection) do not converge if I used the default optimizer or bobyqa. If I fit the model with setting the number of adaptive Gauss-Hermite quadrature points option to 0, the model converges. However, the results are really similar to a GLM without any random factor structure. Therefore, I am not convinced that this is a viable solution and I would like to discuss these convergence problems and the follow-up question how a backward selection process would even be possible under these circumstances.

## Next steps

As I cannot continue with the backward selection process for now, I am trying a forward model selection approach. 


# Forward model selection approach 

```{r, include=TRUE, warning=FALSE}

#I have to drop NAs because anova command can only compare models with exact same number of observations
OCSMCI <- OCSMCI %>% drop_na("ID","group","PictureNaming","Orientation","Semantics","Encoding1","Encoding2","DelayedRecall","DelayedRecallRecognition","EpisodicRecognition","TrailsProcessingSpeed","TrailsExecScore","RuleFinding","RulesLearned","FigureCopy","FigureRecall","Cancellation","Cancellation_FalsPos","InvisibleCancellation","InvisibleCancellation_Revisits")

#I can only add random intercepts and not random slopes: number of observations (=108) < number of random effects (=216) for term (1 + DelayedRecall | ID); the random-effects parameters are probably unidentifiable

# I run 18 models with one subtest as fixed effect each
f1 <- glmer(group ~ 1 + PictureNaming + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f2 <- glmer(group ~ 1 + Orientation + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f3 <- glmer(group ~ 1 + Semantics + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f4 <- glmer(group ~ 1 + Encoding1 + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f5 <- glmer(group ~ 1 + Encoding2 + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f6 <- glmer(group ~ 1 + DelayedRecall + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f7 <- glmer(group ~ 1 + DelayedRecallRecognition + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f8 <- glmer(group ~ 1 + EpisodicRecognition + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f9 <- glmer(group ~ 1 + TrailsProcessingSpeed + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f10 <- glmer(group ~ 1 + TrailsExecScore + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f11 <- glmer(group ~ 1 + RuleFinding + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f12 <- glmer(group ~ 1 + RulesLearned + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f13 <- glmer(group ~ 1 + FigureCopy + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f14 <- glmer(group ~ 1 + FigureRecall + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f15 <- glmer(group ~ 1 + Cancellation + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f16 <- glmer(group ~ 1 + Cancellation_FalsPos + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f17 <- glmer(group ~ 1 + InvisibleCancellation + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f18 <- glmer(group ~ 1 + InvisibleCancellation_Revisits + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)

# model comparison to choose the best model fit and continue with this model
anova(f1,f2,f3,f4,f5,f6,f7,f8,f9,f10,f11,f12,f13,f14,f15,f16,f17,f18)

# I keep the fixed effect from the best model fit and add another subtest as second fixed effect
f201 <- glmer(group ~ 1 + DelayedRecall + PictureNaming + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f202 <- glmer(group ~ 1 + DelayedRecall + Orientation + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f203 <- glmer(group ~ 1 + DelayedRecall + Semantics + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f204 <- glmer(group ~ 1 + DelayedRecall + Encoding1 + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f205 <- glmer(group ~ 1 + DelayedRecall + Encoding2 + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
#f206 <- glmer(group ~ 1 + DelayedRecall + DelayedRecall + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f207 <- glmer(group ~ 1 + DelayedRecall + DelayedRecallRecognition + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f208 <- glmer(group ~ 1 + DelayedRecall + EpisodicRecognition + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f209 <- glmer(group ~ 1 + DelayedRecall + TrailsProcessingSpeed + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f210 <- glmer(group ~ 1 + DelayedRecall + TrailsExecScore + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f211 <- glmer(group ~ 1 + DelayedRecall + RuleFinding + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f212 <- glmer(group ~ 1 + DelayedRecall + RulesLearned + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f213 <- glmer(group ~ 1 + DelayedRecall + FigureCopy + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f214 <- glmer(group ~ 1 + DelayedRecall + FigureRecall + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f215 <- glmer(group ~ 1 + DelayedRecall + Cancellation + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f216 <- glmer(group ~ 1 + DelayedRecall + Cancellation_FalsPos + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f217 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f218 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)

#model comparison
anova(f201,
      f202,
      f203,
      f204,
      f205,
      f207,
      f208,
      f209,
      f210,
      f211,
      f212,
      f213,
      f214,
      f215,
      f216,
      f217,
      f218)

# from now on, this structure will repeat
f301 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + PictureNaming + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f302 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + Orientation + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f303 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + Semantics + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f304 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + Encoding1 + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f305 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + Encoding2 + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f307 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + DelayedRecallRecognition + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f308 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + EpisodicRecognition + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f309 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + TrailsProcessingSpeed + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f310 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + TrailsExecScore + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f311 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + RuleFinding + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f312 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + RulesLearned + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f313 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + FigureCopy + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f314 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + FigureRecall + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f315 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + Cancellation + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f316 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + Cancellation_FalsPos + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f317 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)

anova(f301,
      f302,
      f303,
      f304,
      f305,
      f307,
      f308,
      f309,
      f310,
      f311,
      f312,
      f313,
      f314,
      f315,
      f316,
      f317)

f401 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + PictureNaming + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f402 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + Orientation + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f403 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + Semantics + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f404 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + Encoding1 + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f405 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + Encoding2 + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f407 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + DelayedRecallRecognition + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f408 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + EpisodicRecognition + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f409 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + TrailsProcessingSpeed + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f410 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + TrailsExecScore + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f411 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + RuleFinding + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f412 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + RulesLearned + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f413 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureCopy + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f414 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f415 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + Cancellation + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f416 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + Cancellation_FalsPos + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)

anova(f401,
      f402,
      f403,
      f404,
      f405,
      f407,
      f408,
      f409,
      f410,
      f411,
      f412,
      f413,
      f414,
      f415,
      f416)

f501 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f502 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + Orientation + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f503 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + Semantics + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f504 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + Encoding1 + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f505 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + Encoding2 + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f507 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + DelayedRecallRecognition + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f508 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + EpisodicRecognition + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f509 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + TrailsProcessingSpeed + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f510 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + TrailsExecScore + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f511 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + RuleFinding + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f512 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + RulesLearned + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f513 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + FigureCopy + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f515 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + Cancellation + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f516 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + Cancellation_FalsPos + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)

anova(f501,
      f502,
      f503,
      f504,
      f505,
      f507,
      f508,
      f509,
      f510,
      f511,
      f512,
      f513,
      f515,
      f516)

f602 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + Orientation + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f603 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + Semantics + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f604 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + Encoding1 + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f605 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + Encoding2 + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f607 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + DelayedRecallRecognition + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f608 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + EpisodicRecognition + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f609 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + TrailsProcessingSpeed + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f610 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + TrailsExecScore + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f611 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + RuleFinding + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f612 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + RulesLearned + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f613 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + FigureCopy + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f615 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + Cancellation + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
f616 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + Cancellation_FalsPos + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)

anova(f602,
      f603,
      f604,
      f605,
      f607,
      f608,
      f609,
      f610,
      f611,
      f612,
      f613,
      f615,
      f616)

f702 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + RulesLearned + Orientation + (1|ID),data=OCSMCI, family='binomial')
f703 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + RulesLearned + Semantics + (1|ID),data=OCSMCI, family='binomial')
f704 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + RulesLearned + Encoding1 + (1|ID),data=OCSMCI, family='binomial')
f705 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + RulesLearned + Encoding2 + (1|ID),data=OCSMCI, family='binomial')
f707 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + RulesLearned + DelayedRecallRecognition + (1|ID),data=OCSMCI, family='binomial')
f708 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + RulesLearned + EpisodicRecognition + (1|ID),data=OCSMCI, family='binomial')
f709 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + RulesLearned + TrailsProcessingSpeed + (1|ID),data=OCSMCI, family='binomial')
f710 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + RulesLearned + TrailsExecScore + (1|ID),data=OCSMCI, family='binomial')
f711 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + RulesLearned + RuleFinding + (1|ID),data=OCSMCI, family='binomial')
f713 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + RulesLearned + FigureCopy + (1|ID),data=OCSMCI, family='binomial')
f715 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + RulesLearned + Cancellation + (1|ID),data=OCSMCI, family='binomial')
f716 <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + RulesLearned + Cancellation_FalsPos + (1|ID),data=OCSMCI, family='binomial')

anova(f702,
      f703,
      f704,
      f705,
      f707,
      f708,
      f709,
      f710,
      f711,
      f713,
      f715,
      f716)

f802 <- glmer(group~1+DelayedRecall+InvisibleCancellation_Revisits+InvisibleCancellation+FigureRecall+PictureNaming+RulesLearned+Cancellation_FalsPos+Orientation + (1|ID),data=OCSMCI, family='binomial')
f803 <- glmer(group~1+DelayedRecall+InvisibleCancellation_Revisits+InvisibleCancellation+FigureRecall+PictureNaming+RulesLearned+Cancellation_FalsPos+Semantics + (1|ID),data=OCSMCI, family='binomial')
f804 <- glmer(group~1+DelayedRecall+InvisibleCancellation_Revisits+InvisibleCancellation+FigureRecall+PictureNaming+RulesLearned+Cancellation_FalsPos+Encoding1 + (1|ID),data=OCSMCI, family='binomial')
f805 <- glmer(group~1+DelayedRecall+InvisibleCancellation_Revisits+InvisibleCancellation+FigureRecall+PictureNaming+RulesLearned+Cancellation_FalsPos+Encoding2 + (1|ID),data=OCSMCI, family='binomial')
f807 <- glmer(group~1+DelayedRecall+InvisibleCancellation_Revisits+InvisibleCancellation+FigureRecall+PictureNaming+RulesLearned+Cancellation_FalsPos+DelayedRecallRecognition + (1|ID),data=OCSMCI, family='binomial')
f808 <- glmer(group~1+DelayedRecall+InvisibleCancellation_Revisits+InvisibleCancellation+FigureRecall+PictureNaming+RulesLearned+Cancellation_FalsPos+EpisodicRecognition + (1|ID),data=OCSMCI, family='binomial')
f809 <- glmer(group~1+DelayedRecall+InvisibleCancellation_Revisits+InvisibleCancellation+FigureRecall+PictureNaming+RulesLearned+Cancellation_FalsPos+TrailsProcessingSpeed + (1|ID),data=OCSMCI, family='binomial')
f810 <- glmer(group~1+DelayedRecall+InvisibleCancellation_Revisits+InvisibleCancellation+FigureRecall+PictureNaming+RulesLearned+Cancellation_FalsPos+TrailsExecScore + (1|ID),data=OCSMCI, family='binomial')
f811 <- glmer(group~1+DelayedRecall+InvisibleCancellation_Revisits+InvisibleCancellation+FigureRecall+PictureNaming+RulesLearned+Cancellation_FalsPos+RuleFinding + (1|ID),data=OCSMCI, family='binomial')
f813 <- glmer(group~1+DelayedRecall+InvisibleCancellation_Revisits+InvisibleCancellation+FigureRecall+PictureNaming+RulesLearned+Cancellation_FalsPos+FigureCopy + (1|ID),data=OCSMCI, family='binomial')
f815 <- glmer(group~1+DelayedRecall+InvisibleCancellation_Revisits+InvisibleCancellation+FigureRecall+PictureNaming+RulesLearned+Cancellation_FalsPos+Cancellation + (1|ID),data=OCSMCI, family='binomial')

anova(f802,
      f803,
      f804,
      f805,
      f807,
      f808,
      f809,
      f810,
      f811,
      f813,
      f815)

f902 <- glmer(group~1+DelayedRecall+InvisibleCancellation_Revisits+InvisibleCancellation+FigureRecall+PictureNaming+RulesLearned+Cancellation_FalsPos+Semantics+Orientation + (1|ID),data=OCSMCI, family='binomial')
f904 <- glmer(group~1+DelayedRecall+InvisibleCancellation_Revisits+InvisibleCancellation+FigureRecall+PictureNaming+RulesLearned+Cancellation_FalsPos+Semantics+Encoding1 + (1|ID),data=OCSMCI, family='binomial')
f905 <- glmer(group~1+DelayedRecall+InvisibleCancellation_Revisits+InvisibleCancellation+FigureRecall+PictureNaming+RulesLearned+Cancellation_FalsPos+Semantics+Encoding2 + (1|ID),data=OCSMCI, family='binomial')
f907 <- glmer(group~1+DelayedRecall+InvisibleCancellation_Revisits+InvisibleCancellation+FigureRecall+PictureNaming+RulesLearned+Cancellation_FalsPos+Semantics+DelayedRecallRecognition + (1|ID),data=OCSMCI, family='binomial')
f908 <- glmer(group~1+DelayedRecall+InvisibleCancellation_Revisits+InvisibleCancellation+FigureRecall+PictureNaming+RulesLearned+Cancellation_FalsPos+Semantics+EpisodicRecognition + (1|ID),data=OCSMCI, family='binomial')
f909 <- glmer(group~1+DelayedRecall+InvisibleCancellation_Revisits+InvisibleCancellation+FigureRecall+PictureNaming+RulesLearned+Cancellation_FalsPos+Semantics+TrailsProcessingSpeed + (1|ID),data=OCSMCI, family='binomial')
f910 <- glmer(group~1+DelayedRecall+InvisibleCancellation_Revisits+InvisibleCancellation+FigureRecall+PictureNaming+RulesLearned+Cancellation_FalsPos+Semantics+TrailsExecScore + (1|ID),data=OCSMCI, family='binomial')
f911 <- glmer(group~1+DelayedRecall+InvisibleCancellation_Revisits+InvisibleCancellation+FigureRecall+PictureNaming+RulesLearned+Cancellation_FalsPos+Semantics+RuleFinding + (1|ID),data=OCSMCI, family='binomial')
f913 <- glmer(group~1+DelayedRecall+InvisibleCancellation_Revisits+InvisibleCancellation+FigureRecall+PictureNaming+RulesLearned+Cancellation_FalsPos+Semantics+FigureCopy + (1|ID),data=OCSMCI, family='binomial')
f915 <- glmer(group~1+DelayedRecall+InvisibleCancellation_Revisits+InvisibleCancellation+FigureRecall+PictureNaming+RulesLearned+Cancellation_FalsPos+Semantics+Cancellation + (1|ID),data=OCSMCI, family='binomial')

anova(f902,
      f904,
      f905,
      f907,
      f908,
      f909,
      f910,
      f911,
      f913,
      f915)

#overall comparison, look especially for BIC which accounts for model complexity
anova(f6,f218,f317,f414,f501,f612,f716,f803,f908) #I won't continue here because f501 has the best BIC
summary(f501) #Model failed to converge with max|grad| = 0.253762 (tol = 0.001, component 1); Model is nearly unidentifiable: large eigenvalue ratio - Rescale variables?

```
## Summary

I used a foward model selection approach and always compared the BICs of the resulting models. Model f501 has the best BIC but it failed to converge. How is that possible? Can I even interpret the output of the model then? The error message indicates that I could consider rescaling variables.

## Next steps

I found this advice https://rstudio-pubs-static.s3.amazonaws.com/33653_57fc7b8e5d484c909b615d8633c01d51.html and will try following the suggested options:

1) Rescaling
2) Number of iterations
3) Optimizer


# Rescaling, number of iterations, and different optimizers

```{r, include=TRUE, warning=FALSE}

#follow advice from this website: https://rstudio-pubs-static.s3.amazonaws.com/33653_57fc7b8e5d484c909b615d8633c01d51.html

#descriptives
OCSMCI %>% group_by(group) %>%
  summarise(mean(DelayedRecall), min(DelayedRecall), max(DelayedRecall), mean(InvisibleCancellation_Revisits), min(InvisibleCancellation_Revisits), max(InvisibleCancellation_Revisits),mean(InvisibleCancellation), min(InvisibleCancellation), max(InvisibleCancellation), mean(FigureRecall), min(FigureRecall), max(FigureRecall), mean(PictureNaming), min(PictureNaming), max(PictureNaming))

#scale variables
OCSMCI$DelayedRecall <- scale(OCSMCI$DelayedRecall)
OCSMCI$InvisibleCancellation_Revisits <- scale(OCSMCI$InvisibleCancellation_Revisits)
OCSMCI$InvisibleCancellation <- scale(OCSMCI$InvisibleCancellation)
OCSMCI$FigureRecall <- scale(OCSMCI$FigureRecall)
OCSMCI$PictureNaming <- scale(OCSMCI$PictureNaming)

#model with scaled variables
f501s <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + (1|ID),data=OCSMCI, family='binomial', na.action = na.exclude)
summary(f501s) #scaling advice gone but does not converge

#model with scaled variables and more iterations
f501si <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + (1|ID),data=OCSMCI, family='binomial', control=glmerControl(optCtrl=list(maxfun=2e4)))
summary(f501si) #still does not converge

#model with scaled variables, more iterations, and different optimizer (bobyqa)
f501sio <- glmer(group ~ 1 + DelayedRecall + InvisibleCancellation_Revisits + InvisibleCancellation + FigureRecall + PictureNaming + (1|ID),data=OCSMCI, family='binomial', control=glmerControl(optimizer="bobyqa",optCtrl=list(maxfun=2e5)))
summary(f501sio) #this model converges


```
## Summary

The GLM with scaled variables, the proposed number of iterations, and the bobyqa optimizer converges. However, my main question is whether it makes sense that I scale all variables in the model or should I only scale the truly continuous variables? Also, how do I interpret the output that includes scaled variables? Do I unstandardize estimates for interpretation purposes?




