---
title: "LMM MBody4 Short SMLP Analysis Fixation Duration sec"
author: "Daniel Backhaus"
date: "September 2020"
output: 
    html_document:
        toc: yes
        toc_depth: 1
        number_sections: no
        toc_float: TRUE
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#library(rstudioapi)
#setwd(dirname(getActiveDocumentContext()$path))

rm(list=ls())

set.seed(123)

library(MASS)
library(lmerTest)
library(lme4)
library(lattice) 
library(sjmisc)
library(sjPlot)
library(grid)
library(gridExtra)
library(GGally)
library(tidyverse)
source("../src/mixedDesign.v0.6.3.R")
source("../src/GenDatLMM.R")  # simulation functions
# devtools::install_github("hohenstein/remef")
library(remef)      
source("../src/dotplot.RK.R") 


ginv2 <- function(x) # define a function to make the output nicer
  MASS::fractions(provideDimnames(MASS::ginv(x),base=dimnames(x)[2:1]))

theme_set(theme_bw(base_size=18))
```

# README

Data are from a scene-viewing experiment.  Participants viewed 60 images under 3 different stimulus interaction conditions (1 = silent counting; 2 = point At Last Counted Person (after image presentation; 3 = loud counting with hand gestures)

Variables in `DataImageFix.rda`: fixational events during image presentation 

+ VP -	subject number
+ trial	- trial number
+ Img	image number
+ cond - 	1-"SilentCounting" 2-"PointAtLastCountedPerson" 4-"LoudCounting"
+ fd - fixation duration in ms
+ DV - dependent variable (here fd)

# Load data and set contrasts

```{r, echo=FALSE, message=FALSE, warning=FALSE}

#Filename<-"../data/DataImage.rda"
#load(Filename)#Data.Image
#
#Data.Image.Fix<-Data.Image %>% 
#  filter(MBody==4, Fix==1, eventchange==1, sample>=1, nthfix!=1,revnthfix!=1, #is.na(filterout)) %>% 
#  #filter(nthfix>1 & nthfix < 6) %>% 
#  select(VP,trial,Img,cond,fd) %>% 
#  mutate(DV=fd, DV_org=DV) %>% 
#  mutate(cond=as.factor(cond))
#
#Outfile<-sprintf('../data/DataImageFix.rda')
#save(Data.Image.Fix,file=Outfile)


#load data
Filename<-"../data/DataImageFix.rda"
load(Filename)#Data.Image.Fix
d<-Data.Image.Fix

#add noise
noise<-rnorm(length(d$DV), mean = 0, sd = 1/120*1000)
d <- d %>%
  mutate(noise=noise) %>%
  mutate(DV=abs(DV+noise))

#define conditions
d<-d %>% 
    mutate(FB= case_when(
    cond==1 ~ "SilentCounting",
    cond==2 ~ "PointAtLastCountedPerson",
    cond==3 ~ "LoudCounting"
  )) %>%  
  mutate(FB=as.factor(FB),VP=as.factor(VP),Img=as.factor(Img),DV_log=log(DV), DV_reci= 1/DV) %>% 
  mutate(FB = fct_relevel(FB, "SilentCounting", "PointAtLastCountedPerson", "LoudCounting"))

labelA<-levels(d$FB)[1]
labelB<-levels(d$FB)[2]
labelC<-levels(d$FB)[3]


# Create effect coding for Factor (AB) and Factor (AC)
d <- d %>% mutate(FAB.e =case_when(                             #effect coding
                         FB==labelA ~ -0.5,
                         FB==labelB ~ +0.5,
                         FB==labelC  ~ 0)) %>% 
            mutate(FAC.e =case_when(                            #effect coding
                         FB==labelA ~ -0.5,
                         FB==labelB ~ 0,
                         FB==labelC  ~ +0.5))

#presaved contrast
#contrasts(d$FB) <- contr.treatment(3)


# Create cosum contrasts 
t(Xicost <- fractions(rbind(Hyp1=c(X1= -1/2, X2=+ 1/2, X3=0), 
                            Hyp2=c(X1= -1/2, X2= 0, X3= +1/2))))


(Xcost <- ginv2(Xicost))
colnames(Xcost) <- c("FAB","FAC")
contrasts(d$FB) <- Xcost

#rounded values for some plotting reasons
mat_myC <- round(model.matrix(~ FB, d),digit=5)
d[, 14:16] <- mat_myC
names(d)[14:16] <- c("GM", "FAB", "FAC") 

#add model matrix - mathematical 
mat_myC <- model.matrix(~ FB, d)
d[, 17:19] <- mat_myC
names(d)[17:19] <- c("GM.m", "FAB.m", "FAC.m") 

# Have a look
head(d)
summary(d)
glimpse(d)
```


# Check DV Distribution

```{r}
plot(density(d$DV))
plot(density(d$DV_log))
plot(density(d$DV_reci))

boxcox(DV~ Img, data=d)
```

Transformation required?

0 = log

-1 = reciproc

1 = none

```{r}
d <- d %>% 
  mutate(DV=DV_log)
labelDV<-sprintf("Log Fixation Duration")
```


# Fit LMM

## Question 1

What is `lmer(..., control = lmerControl(calc.derivs = FALSE, optimizer = "bobyqa", optCtrl = list(maxfun = 1e6))` good for ?  Are there pitfalls in usage? Do I need to worry about this with `MixedModels`?

```{r cache=TRUE}

lmm.min <- lmer(DV ~ 1+ FB + (1|VP) + (1|Img), data=d, REML=FALSE)
lmm.min2 <- lmer(DV ~ 1+ FAB.m + FAC.m + (1|VP) + (1|Img), data=d, REML=FALSE, control = lmerControl(calc.derivs = FALSE, optimizer = "bobyqa", optCtrl = list(maxfun = 1e6)))
(anova(lmm.min,lmm.min2))

summary(lmm.min2)
```


## Question 2

When does double bar syntax work and when doesn't work it? Is this equivalent to `MixedModels`? 

```{r cache=TRUE}

#remove correlations
lmm.1a <- lmer(DV ~ 1+ FAB.m + FAC.m + (1+ FAB.m + FAC.m ||VP) + (1+ FAB.m + FAC.m ||Img), data=d, REML=FALSE)

lmm.1b <- lmer(DV ~ 1+ FB + (1+ FB ||VP) + (1+ FB ||Img), data=d, REML=FALSE) #zerocorr doesnt work

(anova(lmm.1a,lmm.1b))
summary(lmm.1a)
summary(lmm.1b)
```

# Model selection

```{r cache=TRUE}
lmm.3a <- lmer(DV ~ 1+ FAB.m + FAC.m + (1 |VP) + (1+ FAC.m ||Img), data=d, REML=FALSE)#con
lmm.3b <- lmer(DV ~ 1+ FAB.m + FAC.m + (0+ FAC.m ||VP) + (1+ FAC.m ||Img), data=d, REML=FALSE)#con
anova(lmm.3a, lmm.3b)

#try to reduce more
lmm.4a <- lmer(DV ~ 1+ FAB.m + FAC.m + (1 |VP) + (1 |Img), data=d, REML=FALSE)#con
anova(lmm.4a,lmm.3a)#ns
anova(lmm.4a,lmm.3b)#ns

#with other optimizer a more complex model is possible
lmm.4b <- lmer(DV ~ 1+ FAB.m + FAC.m + (1 |VP) + (1+ FAB.m +FAC.m ||Img), data=d, REML=FALSE, control = lmerControl(calc.derivs = FALSE, optimizer = "bobyqa", optCtrl = list(maxfun = 1e6)))#con


lmm<-lmm.4a

summary(lmm)
summary(rePCA(lmm))

print(summary(lmm), corr=FALSE)
```

# Significance of fixed effects 

## Question 3

Is a difference with such a small effect size reproducible or random? How to power calculate that? In `MixedModels`?


## Model summary 

```{r fig.width=5, fig.height=5}
labels <- c("FAB", "FAC") 

keep.terms <- names(fixef(lmm)[-1])

## Computing p-values via Wald-statistics approximation (treating t as Wald z).   
tab_model(lmm, terms=keep.terms, pred.labels=labels,
          show.se=TRUE, show.stat=TRUE, show.ci = FALSE, string.se = "SE", 
          show.re.var=FALSE, show.obs=TRUE,
          emph.p = FALSE, dv.labels=labelDV , show.icc = FALSE)
summary(lmm)
```

# Questions - once more
 
1. What is `lmer(..., control = lmerControl(calc.derivs = FALSE, optimizer = "bobyqa", optCtrl = list(maxfun = 1e6))` good for ?  Are there pitfalls in usage? Do I need to worry about this with `MixedModels`?
2. When does double bar syntax work and when doesn't work it? Is this equivalent to `MixedModels`? 
3. Is a difference with this small effectsize reproducible or random? How to power calculate that? In `MixedModels`?

# Appendix

```{r}
sessionInfo()
```
