---
title: "Data Example for smlp2020"
author: "Anna Lena Biel"
date: "1 9 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

Sys.setenv(LANG = "en")
options(scipen=999, digits=7)
```

This example will use a simulated dataset. To simulate the data, we will use the function mixedDesign.

The function mixedDesign can be downloaded from an OSF repository: https://osf.io/7ukf6/.
It allows the researcher to flexibly simulate data from a wide variety of between- and within-subject designs, which are commonly analyzed with (generalized) linear models (ANOVA, ANCOVA, multiple regression analysis) or (generalized) linear mixed-effects models. 
For a detailed description of the function and worked examples, please see Appendix C in Schad et al. (2018) on arxiv:  https://arxiv.org/abs/1807.10451

First, load packages and the function mixedDesign.

```{r load, message=FALSE, warning=FALSE}

# load packages
library(MASS) # load before tidyverse!
library(hypr)
library(lme4) # for mixed effect models
library(lmerOut) # https://bitbucket.org/palday/lmerout/src/master/vignettes/pprint.Rmd
library(effects)
library(emmeans)
library(sjPlot)

library(here)
library(tidyverse)  #plotting, tabulating, etc.
library(psych)
library(Rmisc)
library(DescTools)
library(ez)
library(car)

options(contrasts = c('contr.Sum','contr.Poly'),
        decorate.contr.Sum = "")

# load mixedDesign function for simulating data
source("mixedDesign.v0.6.3.R")

```

For a more general explanation about my typical modelling scenarios and my questions about these, please see the additional PDF document. The following explains the background for this specific simulated dataset.

# Background information on simulated dataset

Imagine that the data came from an experiment where participants received transcranial magnetic stimulation (TMS) while they performed a visual delayed match-to-sample task. TMS was delivered over the right parietal cortex at slower (4 Hz) and faster (7 Hz) theta frequencies or not at all (sham) during the retention interval of the task.

In the task, 2 sets of coloured squares were displayed in both visual hemifields. Prior to presentation of the colour sets, an arrow appeared on the screen to indicate wheter the coloured squares in the left or right hemifield needed to be memorised. The two sets then appeared on the screen, followed by a retention interval. This was followed by a probe display where, again, 2 sets of coloured square were displayed in both visual hemifields. Participants had to indicate whether the set in the cued hemifield had changed. The number of items in each hemifield was 4, 5, or 6. 

As a measure of task performance, K-values for each combination of load and hemisphere for each condition per participant were calculated with the formula: K = (hit rate − false alarms) * set size

Thus, we want to simulate data that includes n=36 participants, which were exposed to the tree conditions (within-subjects):

- Factor SET (number of coloured squares per hemifield) with 3 levels: load 4, load 5, load 6

- Factor STIM (transcranial magnetic stimulation condition) with 3 levels: 4Hz TMS, 7Hz TMS, Sham TMS

- Factor HEMI (Cued hemifield to be memorized) with 2 levels: left cue, right cue

```{r fct_levels}

# define factor names and levels to later label data accordingly
Load_levels <- c("load4", "load5", "load6")
Tms_levels <- c("tms4", "tms7", "tmsS")
Cue_levels <- c("cueL", "cueR")
Subj_levels <- c("S01","S02","S03","S04","S05","S06","S07","S08","S09", "S10",
                 "S11","S12","S13","S14","S15","S16","S17","S18","S19", "S20",
                 "S21","S22","S23","S24","S25","S26","S27","S28","S29", "S30",
                 "S31","S32","S33","S34","S35","S36")

```

# Run Data Simulation 

To simulate the data, we will define the relevant parameters that we need. We will use 2 Matrices to store the parameters for cell means and standard deviations:

1. M  = Matrix with 3x3x2=18 entries, 

2. SD = Matrix with 3x3x2=18 entries, 

both are sorted such that the levels of the first factor vary most slowly, whereas the levels of the later factors vary more quickly, i.e.: 

"load4_tms4_cueL", "load4_tms4_cueR", "load4_tms7_cueL", "load4_tms7_cueR", "load4_tmsS_cueL", "load4_tmsS_cueR", 
"load5_tms4_cueL", "load5_tms4_cueR", "load5_tms7_cueL", "load5_tms7_cueR", "load5_tmsS_cueL", "load5_tmsS_cueR", 
"load6_tms4_cueL", "load6_tms4_cueR", "load6_tms7_cueL", "load6_tms7_cueR", "load6_tmsS_cueL", "load6_tmsS_cueR"

Values for means and standard deviations were derived from previous work (Wolinksi et al. 2018 PlosBiol and Bender et al. 2019 BrainTopography). Based on this, we would mainly expect an interaction effect of STIMxLOAD. Precisely, we would expect that the conditions have the following effetcs on K: 4Hz TMS > Sham TMS > 7Hz TMS for CueL trials, but no such sig. differences for CueR trials. No differences in this pattern depending on set size. 

```{r definitions}

# define means and sds as input for the simulation
M <- matrix(c(1.981, 1.575, 1.189, 1.485, 1.621, 1.55, 
              1.970, 1.561, 1.195, 1.489, 1.626, 1.52, 
              1.989, 1.579, 1.180, 1.475, 1.613, 1.58), 
            nrow=1, ncol=18, byrow=FALSE)
SD <- matrix(c(0.782, 0.943, 0.942, 1.221, 0.906, 0.675, 
               0.682, 0.843, 0.942, 1.021, 0.706, 0.875, 
               0.982, 1.043, 0.942, 0.921, 0.806, 0.975), 
             nrow=1, ncol=18, byrow=FALSE)
R <- 0.8

```

Now, we can run the simulation and look at the resulting data

```{r simdat_raw}

set.seed(321) # set for reproducibility
simdat_raw <- mixedDesign(B=NULL, W=c(3,3,2), n=36, M=M, SD=SD, R=R, empirical=TRUE, long=TRUE) # run the simulation

str(simdat_raw) # look at structure of dataset

head(simdat_raw) # show header

```

# Data preprocessing 

The current dataset contains all the relevant data, but is not pretty yet.

We really need to give the columns and factor levels meaningful names...

```{r namecols}

# Rename columns
names(simdat_raw) <-  c("SUBJ", "SET", "STIM", "HEMI", "K")  

# set all factors and subject ID levels 
simdat <- simdat_raw %>% 
  mutate(SET = factor(SET, labels = Load_levels)) %>%
  mutate(STIM = factor(STIM, labels = Tms_levels)) %>%  
  mutate(HEMI = factor(HEMI, labels = Cue_levels)) %>%
  mutate(SUBJ = factor(SUBJ, labels = Subj_levels))

#here()
write.csv2(simdat, "simdat.csv", row.names=FALSE) # write.csv2 uses a comma for the decimal point and a semicolon for the separator, the Excel convention for CSV files in some Western European locales.

# check data structure and data header again.
str(simdat)

head(simdat)
```

Much better.

# Sanity checks: Explore the data

Usually, we would already do some sanity checks on the single-trial data, e.g. in the logfiles. Now, let's check the simulated dataset. 

```{r Sanitydefineplots, echo=FALSE}

# definitions for all following plots
cbPalette <- c("#E69F00", "#56B4E9","#999999", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

theme_set(theme_linedraw(base_size = 16) +
  theme(axis.text.x = element_text(),
        strip.text.y = element_text(angle=90),
        strip.text = element_text(colour= "black"), # facet text
        strip.background = element_rect(colour="NA", size = 0.05, fill="NA"),
        legend.position="right",
        legend.key.size = unit(2,"line"),
        legend.key.width = unit(2, "line"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()
  ))


Cue_labels <- c(cueL = "left hemifield trials", cueR = "right hemifield trials")

Tms_labels <- c(tms4 = "4 Hz", tms7 = "7 Hz", tmsS = "sham")

Load_labels <- c(load4 = "load 4", load5= "load 5", load6 = "load 6")

```


For example, we could plot the data using a boxplot. Let's follow it up by a single subjects plot and the descriptives as well:


```{r Sanityboxplot, echo=FALSE}

# cueL
ggplot(simdat[simdat$HEMI== "cueL",], aes(x=STIM, y=K, fill=STIM)) +
  geom_hline(yintercept=0) +
  geom_boxplot() +
  ggtitle("left hemifield trials") +
  scale_x_discrete(name="", labels = Tms_labels) +
  scale_y_continuous(name="WM capacity index [K]", limits=c(-3,5), breaks=seq(-5,5,by=1)) +
  scale_color_grey(start = 0.6, end = 0.4) +
  scale_fill_manual(name="", values=cbPalette, labels = Tms_labels) +
  facet_grid(. ~ SET, labeller=labeller(SET = Load_labels)) + 
  guides(fill = FALSE, color = FALSE) 

# cueR
ggplot(simdat[simdat$HEMI== "cueR",], aes(x=STIM, y=K, fill=STIM)) +
  geom_hline(yintercept=0) +
  geom_boxplot() +
  ggtitle("right hemifield trials") +
  scale_x_discrete(name="", labels = Tms_labels) +
  scale_y_continuous(name="WM capacity index [K]", limits=c(-3,5), breaks=seq(-5,5,by=1)) +
  scale_color_grey(start = 0.6, end = 0.4) +
  scale_fill_manual(name="", values=cbPalette, labels = Tms_labels) +
  facet_grid(. ~ SET, labeller=labeller(SET = Load_labels)) + 
  guides(fill = FALSE, color = FALSE)

```


```{r Sanitypointplot, echo=FALSE}

# cueL
ggplot(simdat[simdat$HEMI== "cueL",], aes(x=STIM, y=K, color=STIM, fill=STIM)) +
  geom_hline(yintercept=0) +
  geom_point(size=3, shape=21) +
  geom_line(aes(group = SUBJ)) +
  ggtitle("left hemifield trials") +
  scale_x_discrete(name="", labels = Tms_labels) +
  scale_y_continuous(name="WM capacity index [K]", limits=c(-3,5), breaks=seq(-5,5,by=1)) +
  scale_color_grey(start = 0.6, end = 0.4) +
  scale_fill_manual(name="", values=cbPalette, labels = Tms_labels) +
  facet_grid(. ~ SET, labeller=labeller(SET = Load_labels)) + 
  guides(fill = FALSE, color = FALSE) 

# cueR
ggplot(simdat[simdat$HEMI== "cueR",], aes(x=STIM, y=K, color=STIM, fill=STIM)) +
  geom_hline(yintercept=0) +
  geom_point(size=3, shape=21) +
  geom_line(aes(group = SUBJ)) +
  ggtitle("right hemifield trials") +
  scale_x_discrete(name="", labels = Tms_labels) +
  scale_y_continuous(name="WM capacity index [K]", limits=c(-3,5), breaks=seq(-5,5,by=1)) +
  scale_color_grey(start = 0.6, end = 0.4) +
  scale_fill_manual(name="", values=cbPalette, labels = Tms_labels) +
  facet_grid(. ~ SET, labeller=labeller(SET = Load_labels)) + 
  guides(fill = FALSE, color = FALSE)

#descrsK_DescribeBy <- describeBy(simdat$K, group=list(simdat$Stim, simdat$Cue, simdat$Load), digits=3, mat=TRUE)

(descrs_SummarySEwithin <- summarySEwithin(simdat, measurevar="K", withinvars=c("STIM","HEMI", "SET"),idvar="SUBJ", na.rm=FALSE, conf.interval=.95))



```

# Normality? Plot using a qqplot

```{r qqplotasumptions, echo=FALSE}

## Normality? Plot using a qqplot

# cueL
ggplot(simdat[simdat$HEMI== "cueL",], aes(sample= K, color=STIM)) +
  stat_qq(alpha=.5) + 
  stat_qq_line() +
  ggtitle("left hemifield trials") +
  scale_y_continuous( breaks=seq(0,100,by=2)) +
  scale_x_continuous(name="WM capacity index [K]", breaks=seq(-5,5,by=1)) +
  scale_color_manual(name="", values=cbPalette, labels=Tms_labels) +
  facet_grid(STIM ~ SET, labeller=labeller(SET = Load_labels, STIM=Tms_labels)) + 
  guides(color=FALSE)

# cueR
ggplot(simdat[simdat$HEMI== "cueR",], aes(sample= K, color=STIM)) +
  stat_qq(alpha=.5) + 
  stat_qq_line() +
  ggtitle("right hemifield trials") +
  scale_y_continuous( breaks=seq(0,100,by=2)) +
  scale_x_continuous(name="WM capacity index [K]",  breaks=seq(-5,5,by=1)) +
  scale_color_manual(name="", values=cbPalette, labels=Tms_labels) +
  facet_grid(STIM ~ SET, labeller=labeller(SET = Load_labels, STIM=Tms_labels)) + 
  guides(color=FALSE) 

```

# Hpothesis and contrast coding

Hypothesis: K values should be larger for TMS4 than TMSS and smaller for TMS7 than TMSS, but only in the cueL condition (contralateral to TMS site, i.e. receiving TMS) but not in the cueR condition (ipsilateral to TMS site, i.e. not receiving TMS). This is expected in all SET conditions (load 4, 5 and 6)


Thus, we want to test the following null-hypotheses (H0 for intercept and H0 for contrast(s)) -> input to hypr -> therefore we need the following contrast matrices

```{r contrasts hypr, include=T, echo=T}

sdifC_STIM <- hypr(Intercept=1/3 * tms4 + 1/3 * tmsS + 1/3 * tms7~0, tms4vS=tms4~tmsS, tmsSv7=tmsS~tms7, 
                    levels = c("tms4", "tms7", "tmsS"))
sdifC_STIM # show all

sdifC_HEMI <- hypr(Intercept=1/2*cueL + 1/2*cueR~0, cueLvR=cueL~cueR, 
                    levels = c("cueL", "cueR"))
sdifC_HEMI # show all

sdifC_SET <- hypr(Intercept=1/3*load4 + 1/3*load5 + 1/3*load6~0, load5v4=load5~load4, load6v5=load6~load5, 
                   levels = c("load4", "load5", "load6"))
sdifC_SET # show all

```

Set these contrast matrices

```{r roiR Theta contrasts}

## set these contrasts
contrasts(simdat$STIM) <- contr.hypothesis(sdifC_STIM)
contrasts(simdat$HEMI) <- contr.hypothesis(sdifC_HEMI)
contrasts(simdat$SET) <- contr.hypothesis(sdifC_SET)

# check
# fractions(contrasts(simdat$STIM))
# fractions(contrasts(simdat$HEMI))
# fractions(contrasts(simdat$SET))

```

# Model fit STIM * HEMI * SET with random intercepts and slopes

let's try to fit a model with random intercepts for subject and random slopes

model with random slopes for STIM does not converge 

model with random slope for HEMI is a boundary fit with a singular random effects matrix 

```{r lmerFit_StimHemiSet_rs, include=T}

lmerFit_StimHemiSet_rsSTIM  <- lmer( K ~ 1 + STIM * HEMI * SET + (STIM| SUBJ), 
                          data=simdat, REML = TRUE) # beware! Model does not converge

lmerFit_StimHemiSet_rsHEMI  <- lmer( K ~ 1 + STIM * HEMI * SET + (HEMI| SUBJ), 
                          data=simdat, REML = TRUE) # singular fit, try another optimizer
lmerFit_StimHemiSet_rsHEMI2 <- update(lmerFit_StimHemiSet_rsHEMI, 
                                      control=lmerControl(optimizer="nloptwrap")) # still a singular fit

```

# Model fit STIM * HEMI * SET with random intercepts

random slopes seem tricky, we proably don't have enough data like this. so let's simply fit a model with random intercepts for subject

we want to fit this model: K ~ 1 + STIM * HEMI * SET + (1 | SUBJ)

```{r lmerFit_StimHemiSet, include=T}

# fit reml model (and ml model for model comparison)
lmerFit_StimHemiSet  <- lmer( K ~ 1 + STIM * HEMI * SET + (1| SUBJ), 
                          data=simdat, REML = TRUE)
lmerFit_StimHemiSet_ml <- update(lmerFit_StimHemiSet, REML=FALSE)

# save to file
# later, do this in unix terminal: pandoc yourfile.html -o yourfile.docx (after cd path/to/your/directory)
#pprint(summary(lmerFit_StimHemiSet), "lmerFit_StimHemiSet.tex", type = "latex")
#pprint(summary(lmerFit_StimHemiSet), "lmerFit_StimHemiSet.html", type = "html")

# look at reml model
summary(lmerFit_StimHemiSet)

#vcov(lmerFit_StimHemiSet)

```

## Emmeans table & effect plots plots STIM * HEMI * SET

```{r effect plots lmerFit_StimHemiSet, include=T}

# understand interaction through conditioning
em <- emmeans(lmerFit_StimHemiSet, ~  SET * STIM | HEMI)
em

# understand interaction through visualization 
all_eff <- allEffects(lmerFit_StimHemiSet)
plot(all_eff,
     multiline=TRUE,
     ci.style='auto',
     x.var = 'STIM',
     z.var='HEMI',
     main='Effects as modelled',
     ylab='K')


```


## Fitted values plots STIM * HEMI * SET


```{r RoiL fitted values plots lmerFit_StimHemiSet, include=T}

## observed versus fitted values
plot(lmerFit_StimHemiSet, K ~ fitted(.), abline = c(0,1))

## observed versus fitted values by Subject
plot(lmerFit_StimHemiSet, K ~ fitted(.) | SUBJ, abline = c(0,1))

```

## Model coefficients plots STIM * HEMI * SET

```{r plot_model lmerFit_StimHemiSet, include=T}

theme_set(theme_classic(base_size = 12) +
  theme(#axis.text.x = element_text(angle = 45, hjust = 1),
        panel.spacing = unit(2, "lines"),
        strip.text.y = element_text(angle=0),
        strip.text = element_text(colour= "black"), # facet text
        strip.background = element_rect(colour="NA", size = 0.05, fill="NA"),
        legend.position="right",
        legend.box = "vertical",
        legend.key.size = unit(1.5,"line"),
        legend.key.width = unit(3, "line")
        ))

# Forest-plot of estimates.
plot_model(lmerFit_StimHemiSet, colors = "black", type = "est", transform = NULL, title = "", dot.size = 3,line.size = 1) 


# Plot random effects to look for outliers:
# https://www.jaredknowles.com/journal/2014/5/17/mixed-effects-tutorial-2-fun-with-mermod-objects
re1 <- ranef(lmerFit_StimHemiSet, condVar=TRUE, whichel = "SUBJ")
dotplot(re1)

```

## Residual plots STIM * HEMI * SET

QQ-plot has some tails but could be worse
model plot looks more or less like fluffy cloud
residuals frequency distribution looks symmetrical around zero

```{r residual plots lmerFit_StimHemiSet, include=T}

# frequency distribution of residuals
plot(density(resid(lmerFit_StimHemiSet)))

# QQ-Plot
qqmath(lmerFit_StimHemiSet)#, col = long_RoiL_Theta_Time150$COND, pch = long_RoiL_Theta_Time150$TARG)

# standardized residuals versus fitted values 
plot(lmerFit_StimHemiSet, resid(., scaled=TRUE) ~ fitted(.) , abline = 0)

# Plots fitted values and residuals on a common scale
rfs(lmerFit_StimHemiSet)


```

# Model fit STIM * HEMI 

Since Set was not involved in any interaction, we can maybe fit a simpler model?

we want to fit this model: K ~ 1 + STIM * HEMI + (1 | SUBJ)

```{r lmerFit_StimHemi, include=T}

# fit reml model (and ml model for model comparison)
lmerFit_StimHemi  <- lmer( K ~ 1 + STIM * HEMI + (1| SUBJ), 
                          data=simdat, REML = TRUE)
lmerFit_StimHemi_ml <- update(lmerFit_StimHemi, REML=FALSE)

# look at reml model
summary(lmerFit_StimHemi)

```

## Emmeans table & effect plots plots STIM * HEMI * SET

```{r effect plots lmerFit_StimHemi, include=T}

# understand interaction through conditioning
em <- emmeans(lmerFit_StimHemi, ~  HEMI * STIM)
em

# understand interaction through visualization 
all_eff <- allEffects(lmerFit_StimHemi)
plot(all_eff,
     multiline=TRUE,
     ci.style='auto',
     x.var = 'STIM',
     z.var='HEMI',
     main='Effects as modelled',
     ylab='K')


```


## Fitted values plots STIM * HEMI * SET


```{r RoiL fitted values plots lmerFit_StimHemi, include=T}

## observed versus fitted values
plot(lmerFit_StimHemi, K ~ fitted(.), abline = c(0,1))

## observed versus fitted values by Subject
plot(lmerFit_StimHemi, K ~ fitted(.) | SUBJ, abline = c(0,1))

```

## Model coefficients plots STIM * HEMI * SET

```{r plot_model lmerFit_StimHemi, include=T}

theme_set(theme_classic(base_size = 12) +
  theme(#axis.text.x = element_text(angle = 45, hjust = 1),
        panel.spacing = unit(2, "lines"),
        strip.text.y = element_text(angle=0),
        strip.text = element_text(colour= "black"), # facet text
        strip.background = element_rect(colour="NA", size = 0.05, fill="NA"),
        legend.position="right",
        legend.box = "vertical",
        legend.key.size = unit(1.5,"line"),
        legend.key.width = unit(3, "line")
        ))

# Forest-plot of estimates.
plot_model(lmerFit_StimHemi, colors = "black", type = "est", transform = NULL, title = "", dot.size = 3,line.size = 1) 


# Plot random effects to look for outliers:
# https://www.jaredknowles.com/journal/2014/5/17/mixed-effects-tutorial-2-fun-with-mermod-objects
re1 <- ranef(lmerFit_StimHemi, condVar=TRUE, whichel = "SUBJ")
dotplot(re1)

```

## Residual plots STIM * HEMI * SET

again,
QQ-plot has some tails but could be worse
model plot looks more or less like fluffy cloud
residuals frequency distribution looks symmetrical around zero

```{r residual plots lmerFit_StimHemi, include=T}

# frequency distribution of residuals
plot(density(resid(lmerFit_StimHemi)))

# QQ-Plot
qqmath(lmerFit_StimHemi)#, col = long_RoiL_Theta_Time150$COND, pch = long_RoiL_Theta_Time150$TARG)

# standardized residuals versus fitted values 
plot(lmerFit_StimHemi, resid(., scaled=TRUE) ~ fitted(.) , abline = 0)

# Plots fitted values and residuals on a common scale
rfs(lmerFit_StimHemi)


```





# Model comparison

based on its lower AIC & lower BIC, lmerFit_StimHemi_ml may be selecte, but logLik and deviance don't differ much.

```{r model comp, include=T}

anova(lmerFit_StimHemiSet_ml, lmerFit_StimHemi_ml)


```




















