---
title: "Data Example for smlp2020"
author: "Anna Lena Biel"
date: "1 9 2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

Sys.setenv(LANG = "en")
options(scipen=999, digits=7)
```

This example will use a simulated dataset. To simulate the data, we will use the function mixedDesign.

The function mixedDesign can be downloaded from an OSF repository: https://osf.io/7ukf6/.
It allows the researcher to flexibly simulate data from a wide variety of between- and within-subject designs, which are commonly analyzed with (generalized) linear models (ANOVA, ANCOVA, multiple regression analysis) or (generalized) linear mixed-effects models. 
For a detailed description of the function and worked examples, please see Appendix C in Schad et al. (2018) on arxiv:  https://arxiv.org/abs/1807.10451

First, load packages and the function mixedDesign.

```{r load, message=FALSE, warning=FALSE}
# load packages
library(MASS) # load before tidyverse!
library(hypr)
library(lme4) # for mixed effect models

library(tidyverse)  #plotting, tabulating, etc.
```

**Comments by Reinhold Kliegl (rk)**

# Read simulated data

```{r load, message=FALSE, warning=FALSE}
simdat <- read.csv2("../data/simdat.csv") 
```


# Hpothesis and contrast coding

Hypothesis: K values should be larger for TMS4 than TMSS and smaller for TMS7 than TMSS, but only in the cueL condition (contralateral to TMS site, i.e. receiving TMS) but not in the cueR condition (ipsilateral to TMS site, i.e. not receiving TMS). This is expected in all SET conditions (load 4, 5 and 6)

Thus, we want to test the following null-hypotheses (H0 for intercept and H0 for contrast(s)) -> input to hypr -> therefore we need the following contrast matrices

```{r contrasts hypr, include=T, echo=T}

sdifC_STIM <- hypr(Intercept=1/3 * tms4 + 1/3 * tmsS + 1/3 * tms7~0, tms4vS=tms4~tmsS, tmsSv7=tmsS~tms7, 
                    levels = c("tms4", "tms7", "tmsS"))
sdifC_STIM # show all

sdifC_HEMI <- hypr(Intercept=1/2*cueL + 1/2*cueR~0, cueLvR=cueL~cueR, 
                    levels = c("cueL", "cueR"))
sdifC_HEMI # show all

sdifC_SET <- hypr(Intercept=1/3*load4 + 1/3*load5 + 1/3*load6~0, load5v4=load5~load4, load6v5=load6~load5, 
                   levels = c("load4", "load5", "load6"))
sdifC_SET # show all

```

Set these contrast matrices

```{r roiR Theta contrasts}

## set these contrasts
contrasts(simdat$STIM) <- contr.hypothesis(sdifC_STIM)
contrasts(simdat$HEMI) <- contr.hypothesis(sdifC_HEMI)
contrasts(simdat$SET) <- contr.hypothesis(sdifC_SET)
```

# Model fit STIM * HEMI * SET with random intercepts and slopes

let's try to fit a model with random intercepts for subject and random slopes

model with random slopes for STIM does not converge 

model with random slope for HEMI is a boundary fit with a singular random effects matrix 

## AB

```{r lmerFit_StimHemiSet_rs, include=T}

lmerFit_StimHemiSet_rsSTIM  <- lmer( K ~ 1 + STIM * HEMI * SET + (STIM| SUBJ), 
                          data=simdat, REML = TRUE) # beware! Model does not converge

lmerFit_StimHemiSet_rsHEMI  <- lmer( K ~ 1 + STIM * HEMI * SET + (HEMI| SUBJ), 
                          data=simdat, REML = TRUE) # singular fit, try another optimizer
lmerFit_StimHemiSet_rsHEMI2 <- update(lmerFit_StimHemiSet_rsHEMI, 
                                      control=lmerControl(optimizer="nloptwrap")) # still a singular fit

```

## RK

Let's move to indicator variables.

```{r indicators, include=T}
mm <- model.matrix(~ 1 + STIM * HEMI * SET, dat=simdat)
stm1 <- mm[,2]
stm2 <- mm[,3]
hemi <- mm[,4]
set1 <- mm[,5]
set2 <- mm[,6]
```

The preferred `lmer()` specification is as shown in the next chunk. Convergence issues are most likely false positives.

First model from above.

```{r rsSTIM, include=T}
# replicate lmerFit_StimHemiSet_rsSTIM w/ indicators
base1  <- lmer( K ~ 1 + STIM * HEMI * SET + (1 + stm1 + stm2 | SUBJ), 
                data=simdat, REML = FALSE, control=lmerControl(calc.derivs=FALSE))
summary(rePCA(base1))  # converges, but marginally, actually not supported by data
VarCorr(base1) # CPs are unrealistic

# zero-correlation parameter LMM
zcp1  <- lmer( K ~ 1 + STIM * HEMI * SET + (1 + stm1 + stm2 || SUBJ), 
               data=simdat, REML = FALSE, control=lmerControl(calc.derivs=FALSE))
summary(rePCA(zcp1))  # converges, actually not supported by data
VarCorr(zcp1) 

# parsimonious LMM
prsm1 <- lmer( K ~ 1 + STIM * HEMI * SET + (1  + stm2 || SUBJ), 
               data=simdat, REML = FALSE, control=lmerControl(calc.derivs=FALSE))
summary(rePCA(prsm1))  # supported by data
VarCorr(prsm1) 

# ext w/ CP
prsmx1 <- lmer( K ~ 1 + STIM * HEMI * SET + (1  + stm2 | SUBJ), 
               data=simdat, REML = FALSE, control=lmerControl(calc.derivs=FALSE))
summary(rePCA(prsmx1))  # marginally supported by data
VarCorr(prsmx1)         # overparameterizd CP

anova(prsm1, zcp1) 
```

I would stay with `prms1`.

Second model from above. 

```{r rsHEMI,  include=T}
# replicate lmerFit_StimHemiSet_rsSTIM w/ indicators
base2  <- lmer( K ~ 1 + STIM * HEMI * SET + (1 + hemi | SUBJ), 
                data=simdat, REML = FALSE, control=lmerControl(calc.derivs=FALSE))
summary(rePCA(base2))  # converges, but marginally, actually not supported by data
VarCorr(base2) # CPs are unrealistic

# zero-correlation parameter LMM
zcp2  <- lmer( K ~ 1 + STIM * HEMI * SET + (1 + hemi || SUBJ), 
               data=simdat, REML = FALSE, control=lmerControl(calc.derivs=FALSE))
summary(rePCA(zcp2))  # looks good
VarCorr(zcp2) 

anova(zcp2, base2) 
```

# Appendix

```{r}
sessionInfo()
```













